### 2013-10-02 manuscripts/dpp-msbayes/dpp-msbayes.tex
More editing of dpp ms.

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of evolutionary biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can affect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across effected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the affected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

@@

The software package \msb implements such an approach in an
approximate-Bayesian model-choice framework \citep{Hickerson2006,Huang2011}.
Recently, however, \citet{Oaks2012} found that \msb can be strongly biased
toward spuriously estimating biogeographically interesting models of highly
clustered divergence times across taxa.
The results of their simulation-based assessment of the method suggest that a
combination of broad, uniformly distributed priors on nuisance parameters and a
U-shaped prior on divergence models were causing the bias.
The model implemented in \msb uses a discrete uniform prior over the number of
possible divergence parameters, which \citet{Oaks2012} show imposes a prior on
divergence models that puts most of the prior mass on models with either very
few or very many divergence times shared across the taxa.
Furthermore, uniform priors used for most of the model's nuisance parameters
likely hinder the marginal likelihoods of parameter-rich models, i.e., models
with many divergence time parameters \citep[this is often referred to as
Lindley's paradox;][]{Lindley1957}.
Thus, together, the priors on divergence models and nuisance parameters used in
\msb could create a strong bias toward models with very few divergence time
parameters \citep{Oaks2012}.

Following the recommendations of \citet{Oaks2012} we modify the \msb model to
test whether alternative parameterizations and priors will improve the behavior
of the approximate Bayesian model-choice approach.
We implement a Dirichlet-process prior over all possible models of divergence,
and use gamma and beta probability distributions in place of uniform priors on
many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results show that both the uniform priors on nuisance parameters and
the U-shaped prior on divergence models contribute to the biases of \msb.

@@

We assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
We use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
We let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigator must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed contants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.
Lastly, the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection*{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape-parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate scaling parameters
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identicially
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be
non-zero, the prior term $p(\recombinationRate)$ would be added
to Equation \ref{eq:fullModelCompact}, and the prior would be
$\recombinationRate \sim Gamma(\gammaShape{\recombinationRate},
\gammaScale{\recombinationRate})$.
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence time parameters are in coalescent units relative to the size of
a constant reference population, which we denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}).
    \label{eq:thetaC}
\end{equation}
Given the size of the descendant populations are \iid as
$\descendantTheta{}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units $\myTheta{C}/\mutationRate$
generations, which we denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which we
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{
        \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
        {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be fixed and known according the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior we implement is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
We use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which we denote as $\divTimeIndexVector \sim
\priorDPP{}$.
We take a hierarchical approach and use a prior probability
distribution (i.e., hyperprior) for \concentrationParam.
More specifically, we use a gamma-distributed prior $\concentrationParam =
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the user.

@@

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the only prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides the user with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection*{The priors on nuisance parameters}
Following the recommendations of \citet{Oaks2012}, we have replaced the use of
continuous uniform distributions for priors on many of the
model's parameters (\divTime{}, \ancestralTheta{}, \descendantTheta{1}{},
\descendantTheta{2}{}, \bottleScalar{1}{}, \bottleScalar{2}{},
\recombinationRate, \migrationRate{}) with more flexible parametric
distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
Not only do these priors have nice conjugacy properties, but they also provide
the user with much greater flexibility in expressing prior uncertainty
regarding the parameters of the model.

@@

\subsection*{ABC estimation of the posterior of the model}
\subsubsection*{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameters values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments (these are the same statistics
calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

@@

\subsubsection*{Obtaining an approximate posterior from the prior samples}
We use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and then
it is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the number of desired
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For our analyses, we use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performed very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

@@

\subsection*{Assessing power}
We evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
For all of our power simulations, we used a simulated data structure identical
to that of the empirical dataset of Philippine vertebrates analyzed by
\citet{Oaks2012}, which consists of 22 pairs of populations.
For each of the four models, we generated $2\e6$ samples from the prior.
We then generated pseudo-observed datasets from three series of models.
One series of models, which we denote \powerSeriesOld, is identically
distributed as \modelOld except that the divergence times for each of the 22
pairs of populations are randomly drawn from a series of uniform distributions,
$U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8, 1.0, and
2.0, in \globalcoalunit generations.
A second series of models, \powerSeriesUniform, are identically distributed as
\modelUniform except that the 22 divergence times are randomly drawn from the
same series of uniform priors as above.
The third series of models, \powerSeriesExp, is also identically
distributed as \modelUniform except that the 22 divergence times are randomly
drawn from a series of of exponential distributions: $Exp(17.3205)$,
$Exp(8.6603)$, $Exp(5.7735)$, $Exp(4.33013)$, $Exp(3.4641)$, and $Exp(1.7321)$.
These exponential distributions have the same variance as their uniform
counterparts in the first two series of models.

@@

\subsection*{Validation analyses: Ordered divergence models}
Our results show that the ordered models of divergence, \modelDPPOrdered,
perform similarly to sampling over unordered models (\modelDPP) when the prior
is correct (Figs.\
\labelcref{figValidationAccuracyDPPOrderedDPPOrdered,figValidationModelChoiceDPPOrderedDPPOrdered}).
This suggests that the method is not adversely affected by the large increase
in possible discrete models (from 22 unordered to 4140 ordered models) when
there are 8 pairs of populations.

@@

When the true divergence times are randomly drawn from a series of uniform
priors (\powerSeriesUniform), all of the models poorly estimate
\divTimeDispersion (Figs.\
\labelcref{figPowerAccuracyUniformDPP,figPowerAccuracyUniformUniform,figPowerAccuracyUniformOld,figPowerAccuracyUniformUshaped}).
The \modelDPP and \modelUniform models tend to over-estimate the variance in
divergence times (Figs.\
\labelcref{figPowerAccuracyUniformDPP,figPowerAccuracyUniformUniform}), whereas
the \modelOld model, again, underestimates \divTimeDispersion, estimating
values of zero for most replicates across all the data models of
\powerSeriesUniform (Figure \ref{figPowerAccuracyUniformOld}).
Again, the performance of the \modelUshaped model is intermediate between the
\modelOld and \modelDPP/\modelUniform models (Figure
\ref{figPowerAccuracyUniformUshaped}).
This pattern of results across the four models is very similar when applied
to data simulated under the series of models of \powerSeriesOld (Figs.\
\labelcref{figPowerAccuracyOldDPP,figPowerAccuracyOldUniform,figPowerAccuracyOldOld,figPowerAccuracyOldUshaped}).

>>>>

\section*{Introduction}
Understanding the processes that generate biodiversity and regulate
community assembly is a major goal of evolutionary biology.
Given the dynamic nature of our planet, a major component of this goal is
understanding how geological and climatic processes can affect the
evolutionary history of whole groups of co-distributed species and their
associated microbiota, including pathogens.
Such large-scale processes predict patterns of shared evolutionary history
across affected populations of organisms, e.g., temporally clustered
divergences across co-distributed species.
At recent timescales, such events should leave a signature in the genetic
variation within and among the affected lineages.
Thus, being able to accurately estimate models of shared evolutionary
histories across co-distributed taxa would provide an important
tool for better understanding how regional and global biogeographic
processes affect biodiversity.

@@

The software package \msb implements such an approach in an
approximate-Bayesian model-choice framework \citep{Hickerson2006,Huang2011}.
Recently, however, \citet{Oaks2012} found that \msb can be strongly biased
toward spuriously estimating biogeographically interesting models of highly
clustered divergence times across taxa.
The results of their simulation-based assessment of the method suggest that a
combination of broad, uniformly distributed priors on nuisance parameters and a
U-shaped prior on divergence models are causing the bias.
The model implemented in \msb uses a discrete uniform prior over the number of
possible divergence parameters, which \citet{Oaks2012} show imposes a prior on
divergence models that puts most of the prior mass on models with either very
few or very many divergence times shared across the taxa.
Furthermore, uniform priors used for most of the model's nuisance parameters
likely hinder the marginal likelihoods of parameter-rich models, i.e., models
with many divergence time parameters \citep{Lindley1957}.
Thus, together, the priors on divergence models and nuisance parameters used in
\msb could create a strong bias toward models with very few divergence time
parameters \citep{Oaks2012}.

Following the recommendations of \citet{Oaks2012}, we modify the \msb model to
test whether alternative parameterizations and priors will improve the behavior
of the approximate-Bayesian model-choice approach.
We implement a Dirichlet-process prior over all possible models of divergence,
and use gamma and beta probability distributions in place of uniform priors on
many of the model's parameters.
Our simulation-based analyses show that the new implementation has improved
robustness, accuracy, and power compared to the original model.
Also, our results show that both the uniform priors on nuisance parameters and
the U-shaped prior on divergence models contribute to the biases of \msb.

@@

We assume an investigator is interested in inferring the distribution
of divergence times among \npairs{} pairs of populations.
For each pair $i$, \popSampleSize{i}{} genome copies have been sampled, with
\popSampleSize{1}{i} copies sampled from population 1, and \popSampleSize{2}{i}
sampled from population 2.
From these genomes, let \nloci{i} be the number of DNA sequence loci collected
for population pair $i$, and \nlociTotal be the total number of unique loci
sampled across the \npairs{} pairs of populations.
We use \alignment{i}{j} to represent the multiple sequence alignment of
locus $j$ for population pair $i$.
$\alignmentVector = (\alignment{1}{1}, \ldots,
    \alignment{\npairs{}}{\nloci{\npairs{}}})$
is the full dataset,
i.e., a vector of sequence alignments for all pairs and loci.
We let \geneTree{i}{j} represent the gene tree upon which \alignment{i}{j}
evolved according to fixed \hky substitution model parameters \hkyModel{i}{j}.
The investigator must specify the parameters of all
$\hkyModelVector = (\hkyModel{1}{1}, \ldots,
\hkyModel{\npairs{}}{\nloci{\npairs{}}})$
substitution models by which the alignments evolved along the
$\geneTreeVector = (\geneTree{1}{1}, \ldots,
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
gene trees.
Furthermore, the investigator must specify a vector of fixed constants
$\ploidyScalarVector = (\ploidyScalar{1}{1}, \ldots,
\ploidyScalar{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in ploidy among
loci and/or differences in generation times among population pairs.  Lastly,
the investigator must also specify a vector of fixed constants
$\mutationRateScalarConstantVector = (\mutationRateScalarConstant{1}{1},
\ldots, \mutationRateScalarConstant{\npairs{}}{\nloci{\npairs{}}})$
that scale the population-size parameters for known differences in
mutation rates among loci and/or among taxa.

@@

\subsubsection*{Prior terms of Equation \ref{eq:fullModelCompact}}
\begin{linenomath}
The term $p(\locusRateHetShapeParameter)$ is the prior density
function for the shape-parameter of the gamma-distributed prior on
rate heterogeneity among loci.
This prior is $\locusRateHetShapeParameter \sim U(1, 20)$.
The prior probability of the vector of locus-specific mutation-rate scaling parameters
given \locusRateHetShapeParameter then expands out as a product over
the loci
\begin{equation}
    p(\locusMutationRateScalarVector \given \locusRateHetShapeParameter) =
    \prod_{j=1}^{\nlociTotal}
    p(\locusMutationRateScalar{j} \given \locusRateHetShapeParameter),
    \label{eq:locusRateHetPrior}
\end{equation}
where each \locusMutationRateScalar{} is independently and identically
distributed (\iid) as
$\locusMutationRateScalar{} \sim Gamma(\locusRateHetShapeParameter,
1/\locusRateHetShapeParameter)$.
If the recombination rate \recombinationRate is allowed to be
non-zero, the prior term $p(\recombinationRate)$ would be added
to Equation \ref{eq:fullModelCompact}, and the prior would be
$\recombinationRate \sim Gamma(\gammaShape{\recombinationRate},
\gammaScale{\recombinationRate})$.
\end{linenomath}

@@

\begin{linenomath}
Each \divTime{} within \divTimeVector is \iid as $\divTime{} \sim
Gamma(\gammaShape{\divTime{}}, \gammaScale{\divTime{}})$.
Thus, given the number of unique divergence time classes in
\divTimeIndexVector, this determines the probability of prior term
$p(\divTimeVector \given \divTimeIndexVector)$.
The divergence time parameters are in coalescent units relative to the size of
a constant reference population, which we denote \myTheta{C}, that is equal to
the expectation of the prior on the size of the descendant populations
\begin{equation}
    \myTheta{C} = \mathbb{E}(\descendantTheta{}{}).
    \label{eq:thetaC}
\end{equation}
Given the size of the descendant populations are \iid as
$\descendantTheta{}{} \sim Gamma(\gammaShape{\descendantTheta{}{}},
\gammaScale{\descendantTheta{}{}})$,
this becomes
\begin{equation}
    \myTheta{C} = \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}.
    \label{eq:thetaCGamma}
\end{equation}
The \divTime{} parameters are in units of $\myTheta{C}/\mutationRate$
generations, which we denote as \globalcoalunit generations.
Thus, each \divTime{} within \divTimeVector is proportional to time and can be
converted to the number of generations of the reference population, which we
denote \divTime{G_C}, by assuming a mutation rate and multiplying by the
effective size of the reference population
\begin{equation}
    \divTime{G_C} = \divTime \times \frac{
        \gammaShape{\descendantTheta{}{}}\gammaScale{\descendantTheta{}{}}}
        {\mutationRate}.
    \label{eq:divTimeGenerations}
\end{equation}
Thus, for each of the divergence times in \divTimeVector to be on the same
scale, the relative mutation rates among the pairs of populations are assumed
to be fixed and known according to the user-provided values in
\mutationRateScalarConstantVector.
\end{linenomath}

@@

The second prior we implement is based on the Dirichlet-process, which is a
stochastic process that groups elements into an unknown number of discrete
parameter classes \citep{Ferguson1973,Antoniak1974}.
The process is controlled by the concentration parameter \concentrationParam,
which determines how clustered the process will be.
We use the Dirichlet process to place a prior over all possible ordered
partitions of \npairs{} elements, which we denote as $\divTimeIndexVector \sim
\priorDPP{}$.
We take a hierarchical approach and use a prior probability
distribution (i.e., hyperprior) for \concentrationParam.
More specifically, we use a gamma-distributed prior $\concentrationParam \sim
Gamma(\gammaShape{\concentrationParam}, \gammaScale{\concentrationParam})$,
where \gammaShape{\concentrationParam} and \gammaScale{\concentrationParam} are
specified by the user.

@@

\subsection*{Differences between our model and the original \msb model}
\subsubsection*{The prior on divergence models}
One of the key differences between our model and that of \msb \citep{Huang2011}
is the prior distribution on divergence models.
As discussed in \citet{Oaks2012}, in \msb the prior used for
\divTimeIndexVector is a combination of a discrete uniform prior over the
possible number of divergence events \divTimeNum from 1 to \npairs{} with a
multinomial distribution on the number of times each index of \divTimeVector
appears in \divTimeIndexVector, with the constraint that all \divTime{}
parameters are represented at least once (see Equation 2 of \citet{Oaks2012}).
We denote this prior used by \msb as $\divTimeIndexVector \sim \priorOld$.
\citet{Oaks2012} discuss how placing a uniform prior over the number of
divergence parameters (denoted \divTimeNum here, and as \numt{} in
\citet{Huang2011}) imposes an U-shaped prior over discrete divergence
models (\divTimeIndexVector; see
Figure 5B of \citet{Oaks2012}).
To avoid this, we place priors directly on the sample space of divergence
models, thus eliminating the parameter \numt{} from the model.
We introduce two priors on divergence models:
(1) a prior that is uniform over all unordered divergence models, and
(2) a Dirichlet-process prior on all ordered divergence models.
The latter provides the user with a great deal of flexibility in
expressing their prior beliefs about models of divergence.

@@

\subsubsection*{The priors on nuisance parameters}
Following the recommendations of \citet{Oaks2012}, we have replaced the use of
continuous uniform distributions for priors on many of the
model's parameters (\divTime{}, \ancestralTheta{}, \descendantTheta{1}{},
\descendantTheta{2}{}, \bottleScalar{1}{}, \bottleScalar{2}{},
\recombinationRate, \migrationRate{}) with more flexible parametric
distributions from the exponential family.
We introduce gamma-distributed priors for rate parameters that have a sample
space of all positive real numbers (\divTime{}, \ancestralTheta{},
\descendantTheta{1}{}, \descendantTheta{2}{}, \recombinationRate,
\migrationRate{}), and beta-distributed priors for parameters that are
proportions bounded by zero and one (\bottleScalar{1}{} and
\bottleScalar{2}{}).
These priors provide the user with much greater flexibility in expressing prior
uncertainty regarding the parameters of the model.

@@

\subsection*{ABC estimation of the posterior of the model}
\subsubsection*{Sampling from the prior}
To estimate the approximate posterior of Equation \ref{eq:approxModelCompact},
we use an ABC rejection algorithm.
The first step of this algorithm entails collecting a random sample of
parameter values from the joint prior and their associated summary
statistics.
Each sample is generated by
(1) drawing values of all the model's parameters, which we denote \hpvector{},
from their respective prior distributions;
(2) rescaling the divergence times
$\divTimeMapVector = (\divTimeMap{1}, \ldots, \divTimeMap{\npairs{}})$
from units proportional to time to units proportional to the expected number
of mutations via Equation \ref{eq:divTimeScaled} to get 
$\divTimeScaledVector = (\divTimeScaled{1}{1}, \ldots,
\divTimeScaled{\npairs{}}{\nloci{\npairs{}}})$;
(3) simulating gene trees $\geneTreeVector = (\geneTree{1}{1}, \ldots, 
\geneTree{\npairs{}}{\nloci{\npairs{}}})$
for each locus of each population pair by drawing coalescent times from
a multi-population Kingman-coalescent model given the demographic parameters;
(4) simulating sequence alignments 
$\alignmentVector = (\alignment{1}{1}, \ldots, \alignment{\npairs{}}{\nloci{\npairs{}}})$
along the gene trees under the
\hky substitution parameters
$\hkyModelVector = (\hkyModel{1}{1}, \ldots, \hkyModel{\npairs{}}{\nloci{\npairs{}}})$
that have the same number of sequences and sequence lengths as the observed
dataset;
(5) calculating population genetic summary statistics
$\ssVector{} = (\alignmentSS{1}{1}, \ldots, \alignmentSS{\npairs{}}{\nloci{\npairs{}}})$
from the simulated sequence alignments;
% (these are the same statistics
% calculated from the observed alignments, \ssVectorObs{});
and (6) reducing the summary statistics to the means across loci for each
population pair to get
$\ssVector{} = (\alignmentSS{1}{}, \ldots, \alignmentSS{\npairs{}}{})$, which is
the same summary statistic vector estimated from the observed data \ssVectorObs.
After repeating this procedure \numPriorSamples times, we obtain a random
sample of parameter vectors
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$
from the model prior and their associated vectors of summary statistics
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$.

@@

\subsubsection*{Obtaining an approximate posterior from the prior samples}
We use a rejection algorithm to retain an approximate posterior sample of
\paramSampleVector{} from the prior sample
$\paramSampleMatrix = (\paramSampleVector{1}, \ldots, \paramSampleVector{\numPriorSamples})$.
First, the observed summary statistics \ssVectorObs, and the summary statistics
of the prior samples 
$\ssMatrix = (\ssVector{1}, \ldots, \ssVector{\numPriorSamples})$,
are standardized using the means and standard deviations of the statistics from
the prior sample (i.e., the prior mean is subtracted from each statistic, and the
difference is divided by the prior standard deviation).
After all statistics are standardized, the Euclidean distance between
\ssVectorObs and each \ssVector{} within \ssMatrix is calculated.
The samples that fall within a range of tolerance \tol around \ssVectorObs
are retained.
The range of tolerance is determined by specifying the number of desired
posterior samples to be retained.
Post-hoc adjustment of the posterior sample can also be performed with a number
of regression techniques \citep{Beaumont2002,Blum2009,Leuenberger2010}.
For our analyses, we use the general linear model (GLM) regression adjustment
\citep{Leuenberger2010} as implemented in \abctoolbox
\citep[v1.1;][]{ABCtoolbox}, which \citet{Oaks2012} showed performed very
similarly to weighted local-linear regression and multinomial logistic
regression adjustments \citep{Beaumont2002} for \msb posteriors.

@@

\subsection*{Assessing power}
We evaluated the power of the same four models (Table~\ref{tabPriors}) to
detect random variation in divergence times using methods similar to
\citet{Oaks2012}.
For all of our power simulations, we used a simulated data structure identical
to that of the empirical dataset of Philippine vertebrates analyzed by
\citet{Oaks2012}, which consists of 22 pairs of populations.
For each of the four models, we generated $2\e6$ samples from the prior.
We then generated pseudo-observed datasets from three series of models.
One series of models, which we denote \powerSeriesOld, is identically
distributed as \modelOld except that the divergence times for each of the 22
pairs of populations are randomly drawn from a series of uniform distributions,
$U(0, \divt{max})$, where \divt{max} was set to: 0.2, 0.4, 0.6, 0.8, 1.0, and
2.0, in \globalcoalunit generations.
A second series of models, \powerSeriesUniform, are identically distributed as
\modelUniform except that the 22 divergence times are randomly drawn from the
same series of uniform priors as above.
The third series of models, \powerSeriesExp, is also identically
distributed as \modelUniform except that the 22 divergence times are randomly
drawn from a series of of exponential distributions: $Exp(17.3205)$,
$Exp(8.6603)$, $Exp(5.7735)$, $Exp(4.3301)$, $Exp(3.4641)$, and $Exp(1.7321)$.
These exponential distributions have the same variance as their uniform
counterparts in the first two series of models.

@@

\subsection*{Validation analyses: Ordered divergence models}
Our results show that the ordered models of divergence, \modelDPPOrdered,
perform similarly to sampling over unordered models (\modelDPP) when the prior
is correct (Figs.\
\labelcref{figValidationAccuracyDPPOrderedDPPOrdered,figValidationModelChoiceDPPOrderedDPPOrdered}).
This suggests that the method is not adversely affected by the large increase
in the number of possible discrete models (from 22 unordered to 4140 ordered
models) when there are eight pairs of populations.

@@

When the true divergence times are randomly drawn from a series of uniform
priors (\powerSeriesUniform), all of the models poorly estimate
\divTimeDispersion (Figs.\
\labelcref{figPowerAccuracyUniformDPP,figPowerAccuracyUniformUniform,figPowerAccuracyUniformOld,figPowerAccuracyUniformUshaped}).
The \modelDPP and \modelUniform models tend to over-estimate the variance in
divergence times (Figs.\
\labelcref{figPowerAccuracyUniformDPP,figPowerAccuracyUniformUniform}), whereas
the \modelOld model, again, underestimates \divTimeDispersion, estimating
values of zero for most replicates across all the data models of
\powerSeriesUniform (Figure \ref{figPowerAccuracyUniformOld}).
Again, the performance of the \modelUshaped model is intermediate between the
\modelOld and \modelDPP/\modelUniform models (Figure
\ref{figPowerAccuracyUniformUshaped}).
This pattern of results across the four models is very similar to their
behavior when applied to data simulated under the series of models of
\powerSeriesOld (Figs.\
\labelcref{figPowerAccuracyOldDPP,figPowerAccuracyOldUniform,figPowerAccuracyOldOld,figPowerAccuracyOldUshaped}).
